Plant trait cleaning and combining

### 2024-05-26 code/L1/L1_LastStep_DiversityInputObjects.Rmd
add packages, update presence-absence matrix function, run function for each resolution, add outputs from each resolution

# Load required packages
```{r}
library(letsR)
library(mFD)
library(vegan)
library(rnaturalearth)
library(sf)
library(raster)
library(fasterize)
library(dplyr)
library(funbiogeo)
```

@@

Use unprojected data because projected data is causing issues
```{r}
x <- TropicalAndes_plant_occ_forest$decimalLongitude
y <- TropicalAndes_plant_occ_forest$decimalLatitude
plant_xy <- cbind(x, y)
plant_species <- TropicalAndes_plant_occ_forest$species
plants_PAM <- letsR::lets.presab.points(plant_xy, plant_species, xmn = -85, xmx = -60, ymn = -24, ymx = 14, count = TRUE)
summary(plants_PAM)
plot(plants_PAM, xlab = "Longitude", ylab = "Latitude", main = "Plant richness map")
```

```{r}
x <- TropicalAndes_frugivore_occ_forest$decimalLongitude
y <- TropicalAndes_frugivore_occ_forest$decimalLatitude
frugivore_xy <- cbind(x, y)
frugivore_species <- TropicalAndes_frugivore_occ_forest$species
frugivore_PAM <- letsR::lets.presab.points(frugivore_xy, frugivore_species, xmn = -85, xmx = -60, ymn = -24, ymx = 14, count = TRUE)
summary(frugivore_PAM)
plot(frugivore_PAM, xlab = "Longitude", ylab = "Latitude", main = "Frugivore richness map")
```

## Turn PAM into matrix
```{r}
plants_PAM_matrix <- lets.presab.points(plant_xy, plant_species, xmn = -85, xmx = -60, ymn = -24, ymx = 14, show.matrix = TRUE, remove.cells = TRUE, count = TRUE)
frugivore_PAM_matrix <- lets.presab.points(frugivore_xy, frugivore_species, xmn = -85, xmx = -60, ymn = -24, ymx = 14, show.matrix = TRUE, remove.cells = TRUE, count = TRUE)
```

## Add rownames
```{r}
# Get the number of rows in the matrix
num_rows_plant <- nrow(plants_PAM_matrix)

# Generate unique row names
row_names_plant <- paste0("cell", 1:num_rows_plant)

# Assign row names to the matrix
rownames(plants_PAM_matrix) <- row_names_plant

column_names_plant <- colnames(plants_PAM_matrix)
```

```{r}
# Get the number of rows in the matrix
num_rows_frugivore <- nrow(frugivore_PAM_matrix)

# Generate unique row names
row_names_frugivore <- paste0("cell", 1:num_rows_frugivore)

# Assign row names to the matrix
rownames(frugivore_PAM_matrix) <- row_names_frugivore

column_names_frugivore <- colnames(frugivore_PAM_matrix)
```

## Turn data into correct data types for inputs into the trait categories dataframe
```{r}
plant_traits$Growth.Form <- as.factor(plant_traits$Growth.Form)
plant_traits$Plant.Height_m <- as.numeric(plant_traits$Plant.Height_m)
plant_traits$Fruit.Length_mm <- as.numeric(plant_traits$Fruit.Length_mm)
plant_traits$Fruit.Type <- as.factor(plant_traits$Fruit.Type)
```

```{r}
frugivore_traits$diet_cat <- as.factor(frugivore_traits$diet_cat)
frugivore_traits$body_mass_e <- as.numeric(frugivore_traits$body_mass_e)
frugivore_traits$body_size_mm <- as.numeric(frugivore_traits$body_size_mm)
frugivore_traits$generation_time <- as.numeric(frugivore_traits$generation_time)
```

## Remove the species from PAM that have no occurrences
```{r}
# Remove columns with sum equal to zero
PAM_plant_site_final <- plants_PAM_matrix[, colSums(plants_PAM_matrix) != 0]
PAM_frugivore_site_final <- frugivore_PAM_matrix[, colSums(frugivore_PAM_matrix) != 0]

#Save coordinates for later
site_loc_key_plant <- PAM_plant_site_final[,1:2]
site_loc_key_frugivore <- PAM_frugivore_site_final[,1:2]

PAM_plant_site_final <- PAM_plant_site_final[,-c(1:2)]
PAM_frugivore_site_final <- PAM_frugivore_site_final[,-(1:2)]

colnames_plant <- colnames(PAM_plant_site_final)
colnames_frugivore <- colnames(PAM_frugivore_site_final)
```

```{r}
str(PAM_plant_site_final)
str(PAM_frugivore_site_final)
```

## Remove species names from trait matrix not in the PAM
```{r}
plant_traits_df_subset <- plant_traits %>%
                            filter(species %in% colnames_plant) %>%
                            distinct(species, .keep_all = TRUE)
frugivore_traits_df_subset <- frugivore_traits %>%
                                filter(IUCN_species_name %in% colnames_frugivore) %>%
                                distinct(IUCN_species_name, .keep_all = TRUE)

@@

# Export data
```{r}
# objects to save
## sf objects
saveRDS(plants_sf_species, file = file.path(data_path_L1,"plants_sf_species.rds"))
saveRDS(frugivores_sf_species, file = file.path(data_path_L1,"frugivores_sf_species.rds"))
saveRDS(Americas, file = file.path(data_path_L1, "Americas.rds"))
saveRDS(TApoly, file = file.path(data_path_L1,"TApoly.rds"))
saveRDS(TropicalAndes_IUCNHabitat_Forest, file = file.path(data_path_L1,"TropicalAndes_IUCNHabitat_Forest.rds"))
```

```{r}
# PAM objects
saveRDS(plants_PAM, file = file.path(data_path_L1,"plants_PAM.rds"))
saveRDS(frugivore_PAM, file = file.path(data_path_L1,"frugivore_PAM.rds"))
```

```{r}
# objects for functional diversity
saveRDS(site_loc_key_plant, file = file.path(data_path_L1,"site_loc_key_plant.rds"))
saveRDS(site_loc_key_frugivore, file = file.path(data_path_L1,"site_loc_key_frugivore.rds"))
saveRDS(PAM_plant_site_final, file = file.path(data_path_L1,"PAM_plant_site_final.rds"))
saveRDS(PAM_frugivore_site_final, file = file.path(data_path_L1,"PAM_frugivore_site_final.rds"))
saveRDS(plant_traits_df_subset, file = file.path(data_path_L1,"plant_traits_df_final.rds"))
saveRDS(frugivore_traits_df_subset, file = file.path(data_path_L1,"frugivore_traits_df_final.rds"))
```

>>>>

# Load required packages
```{r}
library(letsR)
library(mFD)
library(vegan)
library(rnaturalearth)
library(sf)
library(raster)
library(fasterize)
library(funbiogeo)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
```

@@

```{r}
# Define the function
create_presence_absence_matrix <- function(resolution_meters, species_sf) {
  # Make Grid
  TAGrid <- TApoly %>%
    st_make_grid(cellsize = c(resolution_meters)) %>%
    st_intersection(TropicalAndes_IUCNHabitat_Forest) %>%
    st_cast("MULTIPOLYGON") %>%
    st_sf() %>%
    mutate(cellid = row_number(),
           centroid = st_centroid(geometry)) %>%
    mutate(longitude = st_coordinates(centroid)[, 1],
           latitude = st_coordinates(centroid)[, 2]) %>%
    dplyr::select(-centroid)

  # Join with species data and filter out NA species
  species_grid <- TAGrid %>%
    st_join(species_sf) %>%
    filter(!is.na(species)) %>%
    dplyr::select(cellid, longitude, latitude, species)

  # Create Presence-Absence Matrix
  presence_absence_matrix <- species_grid %>%
    st_set_geometry(NULL) %>%
    mutate(presence = 1) %>%
    pivot_wider(names_from = species, values_from = presence, values_fill = list(presence = 0)) %>%
    arrange(cellid) %>%
    dplyr::select(cellid, longitude, latitude, everything())

  # Convert to matrix and set row names
  presence_absence_matrix <- as.data.frame(presence_absence_matrix)
  rownames(presence_absence_matrix) <- presence_absence_matrix$cellid
  presence_absence_matrix <- presence_absence_matrix %>%
    dplyr::select(-cellid) %>%
    as.matrix()

  return(presence_absence_matrix)
}
```

## 100 km 
```{r}
plant_PAM_100km <- create_presence_absence_matrix(100000, plants_sf_species)
frugivore_PAM_100km <- create_presence_absence_matrix(100000, frugivores_sf_species)
```

```{r}
#check str 
plant_PAM_100km[1:4, 1:4]
frugivore_PAM_100km[1:4, 1:4]
```

### Remove the species from PAM that have no occurrences
```{r}
# Remove columns with sum equal to zero
PAM_plant_site_final_100km <- plant_PAM_100km[, colSums(plant_PAM_100km) != 0]
PAM_frugivore_site_final_100km <- frugivore_PAM_100km[, colSums(frugivore_PAM_100km) != 0]

#Save coordinates for later
site_loc_key_plant_100km <- PAM_plant_site_final_100km[,1:2]
site_loc_key_frugivore_100km <- PAM_frugivore_site_final_100km[,1:2]

PAM_plant_site_final_100km <- PAM_plant_site_final_100km[,-c(1:2)]
PAM_frugivore_site_final_100km <- PAM_frugivore_site_final_100km[,-(1:2)]

colnames_plant_100km <- colnames(PAM_plant_site_final_100km)
colnames_frugivore_100km <- colnames(PAM_frugivore_site_final_100km)
```

```{r}
str(PAM_plant_site_final_100km)
str(PAM_frugivore_site_final_100km)
```
## 75 km 
```{r}
plant_PAM_75km <- create_presence_absence_matrix(75000, plants_sf_species)
frugivore_PAM_75km <- create_presence_absence_matrix(75000, frugivores_sf_species)
```

```{r}
#check str 
plant_PAM_75km[1:4, 1:4]
frugivore_PAM_75km[1:4, 1:4]
```

### Remove the species from PAM that have no occurrences
```{r}
# Remove columns with sum equal to zero
PAM_plant_site_final_75km <- plant_PAM_75km[, colSums(plant_PAM_75km) != 0]
PAM_frugivore_site_final_75km <- frugivore_PAM_75km[, colSums(frugivore_PAM_75km) != 0]

#Save coordinates for later
site_loc_key_plant_75km <- PAM_plant_site_final_75km[,1:2]
site_loc_key_frugivore_75km <- PAM_frugivore_site_final_75km[,1:2]

PAM_plant_site_final_75km <- PAM_plant_site_final_75km[,-c(1:2)]
PAM_frugivore_site_final_75km <- PAM_frugivore_site_final_75km[,-(1:2)]

colnames_plant_75km <- colnames(PAM_plant_site_final_75km)
colnames_frugivore_75km <- colnames(PAM_frugivore_site_final_75km)
```

```{r}
str(PAM_plant_site_final_75km)
str(PAM_frugivore_site_final_75km)
```

## 50 km 
```{r}
plant_PAM_50km <- create_presence_absence_matrix(50000, plants_sf_species)
frugivore_PAM_50km <- create_presence_absence_matrix(50000, frugivores_sf_species)
```

```{r}
#check str 
plant_PAM_50km[1:4, 1:4]
frugivore_PAM_50km[1:4, 1:4]
```

### Remove the species from PAM that have no occurrences
```{r}
# Remove columns with sum equal to zero
PAM_plant_site_final_50km <- plant_PAM_50km[, colSums(plant_PAM_50km) != 0]
PAM_frugivore_site_final_50km <- frugivore_PAM_50km[, colSums(frugivore_PAM_50km) != 0]

#Save coordinates for later
site_loc_key_plant_50km <- PAM_plant_site_final_50km[,1:2]
site_loc_key_frugivore_50km <- PAM_frugivore_site_final_50km[,1:2]

PAM_plant_site_final_50km <- PAM_plant_site_final_50km[,-c(1:2)]
PAM_frugivore_site_final_50km <- PAM_frugivore_site_final_50km[,-(1:2)]

colnames_plant_50km <- colnames(PAM_plant_site_final_50km)
colnames_frugivore_50km <- colnames(PAM_frugivore_site_final_50km)
```

```{r}
str(PAM_plant_site_final_50km)
str(PAM_frugivore_site_final_50km)
```

## 25 km 
```{r}
plant_PAM_25km <- create_presence_absence_matrix(25000, plants_sf_species)
frugivore_PAM_25km <- create_presence_absence_matrix(25000, frugivores_sf_species)
```

```{r}
#check str 
plant_PAM_25km[1:4, 1:4]
frugivore_PAM_25km[1:4, 1:4]
```

### Remove the species from PAM that have no occurrences
```{r}
# Remove columns with sum equal to zero
PAM_plant_site_final_25km <- plant_PAM_25km[, colSums(plant_PAM_25km) != 0]
PAM_frugivore_site_final_25km <- frugivore_PAM_25km[, colSums(frugivore_PAM_25km) != 0]

#Save coordinates for later
site_loc_key_plant_25km <- PAM_plant_site_final_25km[,1:2]
site_loc_key_frugivore_25km <- PAM_frugivore_site_final_25km[,1:2]

PAM_plant_site_final_25km <- PAM_plant_site_final_25km[,-c(1:2)]
PAM_frugivore_site_final_25km <- PAM_frugivore_site_final_25km[,-(1:2)]

colnames_plant_25km <- colnames(PAM_plant_site_final_25km)
colnames_frugivore_25km <- colnames(PAM_frugivore_site_final_25km)
```

```{r}
str(PAM_plant_site_final_25km)
str(PAM_frugivore_site_final_25km)
```

## 10 km 
```{r}
plant_PAM_10km <- create_presence_absence_matrix(10000, plants_sf_species)
frugivore_PAM_10km <- create_presence_absence_matrix(10000, frugivores_sf_species)
```

```{r}
#check str 
plant_PAM_10km[1:4, 1:4]
frugivore_PAM_10km[1:4, 1:4]
```

### Remove the species from PAM that have no occurrences
```{r}
# Remove columns with sum equal to zero
PAM_plant_site_final_10km <- plant_PAM_10km[, colSums(plant_PAM_10km) != 0]
PAM_frugivore_site_final_10km <- frugivore_PAM_10km[, colSums(frugivore_PAM_10km) != 0]

#Save coordinates for later
site_loc_key_plant_10km <- PAM_plant_site_final_10km[,1:2]
site_loc_key_frugivore_10km <- PAM_frugivore_site_final_10km[,1:2]

PAM_plant_site_final_10km <- PAM_plant_site_final_10km[,-c(1:2)]
PAM_frugivore_site_final_10km <- PAM_frugivore_site_final_10km[,-(1:2)]

colnames_plant_10km <- colnames(PAM_plant_site_final_10km)
colnames_frugivore_10km <- colnames(PAM_frugivore_site_final_10km)
```

```{r}
str(PAM_plant_site_final_10km)
str(PAM_frugivore_site_final_10km)
```

## 5 km 
```{r}
plant_PAM_5km <- create_presence_absence_matrix(5000, plants_sf_species)
frugivore_PAM_5km <- create_presence_absence_matrix(5000, frugivores_sf_species)
```

```{r}
#check str 
plant_PAM_5km[1:4, 1:4]
frugivore_PAM_5km[1:4, 1:4]
```

### Remove the species from PAM that have no occurrences
```{r}
# Remove columns with sum equal to zero
PAM_plant_site_final_5km <- plant_PAM_5km[, colSums(plant_PAM_5km) != 0]
PAM_frugivore_site_final_5km <- frugivore_PAM_5km[, colSums(frugivore_PAM_5km) != 0]

#Save coordinates for later
site_loc_key_plant_5km <- PAM_plant_site_final_5km[,1:2]
site_loc_key_frugivore_5km <- PAM_frugivore_site_final_5km[,1:2]

PAM_plant_site_final_5km <- PAM_plant_site_final_5km[,-c(1:2)]
PAM_frugivore_site_final_5km <- PAM_frugivore_site_final_5km[,-(1:2)]

colnames_plant_5km <- colnames(PAM_plant_site_final_5km)
colnames_frugivore_5km <- colnames(PAM_frugivore_site_final_5km)
```

```{r}
str(PAM_plant_site_final_5km)
str(PAM_frugivore_site_final_5km)
```

## 1 km 
```{r}
plant_PAM_1km <- create_presence_absence_matrix(1000, plants_sf_species)
frugivore_PAM_1km <- create_presence_absence_matrix(1000, frugivores_sf_species)
```

```{r}
#check str 
plant_PAM_1km[1:4, 1:4]
frugivore_PAM_1km[1:4, 1:4]
```

### Remove the species from PAM that have no occurrences
```{r}
# Remove columns with sum equal to zero
PAM_plant_site_final_1km <- plant_PAM_1km[, colSums(plant_PAM_1km) != 0]
PAM_frugivore_site_final_1km <- frugivore_PAM_1km[, colSums(frugivore_PAM_1km) != 0]

#Save coordinates for later
site_loc_key_plant_1km <- PAM_plant_site_final_1km[,1:2]
site_loc_key_frugivore_1km <- PAM_frugivore_site_final_1km[,1:2]

PAM_plant_site_final_1km <- PAM_plant_site_final_1km[,-c(1:2)]
PAM_frugivore_site_final_1km <- PAM_frugivore_site_final_1km[,-(1:2)]

colnames_plant_1km <- colnames(PAM_plant_site_final_1km)
colnames_frugivore_1km <- colnames(PAM_frugivore_site_final_1km)
```

```{r}
str(PAM_plant_site_final_1km)
str(PAM_frugivore_site_final_1km)
```

## Remove species names from trait matrix not in the PAM
```{r}
plant_traits_df_subset <- plant_traits %>%
                            filter(species %in% colnames_plant_100km) %>%
                            distinct(species, .keep_all = TRUE)
frugivore_traits_df_subset <- frugivore_traits %>%
                                filter(IUCN_species_name %in% colnames_frugivore_100km) %>%
                                distinct(IUCN_species_name, .keep_all = TRUE)

@@

# Export data
## sf objects
```{r}
saveRDS(plants_sf_species, file = file.path(data_path_L1,"plants_sf_species.rds"))
saveRDS(frugivores_sf_species, file = file.path(data_path_L1,"frugivores_sf_species.rds"))
saveRDS(Americas, file = file.path(data_path_L1, "Americas.rds"))
saveRDS(TApoly, file = file.path(data_path_L1,"TApoly.rds"))
saveRDS(TropicalAndes_IUCNHabitat_Forest, file = file.path(data_path_L1,"TropicalAndes_IUCNHabitat_Forest.rds"))
```

# PAM objects
```{r}
saveRDS(plant_PAM_100km, file = file.path(data_path_L1,"plant_PAM_100km.rds"))
saveRDS(frugivore_PAM_100km, file = file.path(data_path_L1,"frugivore_PAM_100km.rds"))
```

```{r}
saveRDS(plant_PAM_75km, file = file.path(data_path_L1,"plant_PAM_75km.rds"))
saveRDS(frugivore_PAM_75km, file = file.path(data_path_L1,"frugivore_PAM_75km.rds"))
```

```{r}
saveRDS(plant_PAM_50km, file = file.path(data_path_L1,"plant_PAM_50km.rds"))
saveRDS(frugivore_PAM_50km, file = file.path(data_path_L1,"frugivore_PAM_50km.rds"))
```

```{r}
saveRDS(plant_PAM_25km, file = file.path(data_path_L1,"plant_PAM_25km.rds"))
saveRDS(frugivore_PAM_25km, file = file.path(data_path_L1,"frugivore_PAM_25km.rds"))
```

```{r}
saveRDS(plant_PAM_10km, file = file.path(data_path_L1,"plant_PAM_10km.rds"))
saveRDS(frugivore_PAM_10km, file = file.path(data_path_L1,"frugivore_PAM_10km.rds"))
```

```{r}
saveRDS(plant_PAM_5km, file = file.path(data_path_L1,"plant_PAM_5km.rds"))
saveRDS(frugivore_PAM_5km, file = file.path(data_path_L1,"frugivore_PAM_5km.rds"))
```

```{r}
saveRDS(plant_PAM_1km, file = file.path(data_path_L1,"plant_PAM_1km.rds"))
saveRDS(frugivore_PAM_1km, file = file.path(data_path_L1,"frugivore_PAM_1km.rds"))
```

## objects for functional diversity
```{r}
saveRDS(plant_traits_df_subset, file = file.path(data_path_L1,"plant_traits_df_final.rds"))
saveRDS(frugivore_traits_df_subset, file = file.path(data_path_L1,"frugivore_traits_df_final.rds"))
```

```{r}
saveRDS(site_loc_key_plant_100km, file = file.path(data_path_L1,"site_loc_key_plant_100km.rds"))
saveRDS(site_loc_key_frugivore_100km, file = file.path(data_path_L1,"site_loc_key_frugivore_100km.rds"))
saveRDS(PAM_plant_site_final_100km, file = file.path(data_path_L1,"PAM_plant_site_final_100km.rds"))
saveRDS(PAM_frugivore_site_final_100km, file = file.path(data_path_L1,"PAM_frugivore_site_final_100km.rds"))
```

```{r}
saveRDS(site_loc_key_plant_75km, file = file.path(data_path_L1,"site_loc_key_plant_75km.rds"))
saveRDS(site_loc_key_frugivore_75km, file = file.path(data_path_L1,"site_loc_key_frugivore_75km.rds"))
saveRDS(PAM_plant_site_final_75km, file = file.path(data_path_L1,"PAM_plant_site_final_75km.rds"))
saveRDS(PAM_frugivore_site_final_75km, file = file.path(data_path_L1,"PAM_frugivore_site_final_75km.rds"))
```

```{r}
saveRDS(site_loc_key_plant_50km, file = file.path(data_path_L1,"site_loc_key_plant_50km.rds"))
saveRDS(site_loc_key_frugivore_50km, file = file.path(data_path_L1,"site_loc_key_frugivore_50km.rds"))
saveRDS(PAM_plant_site_final_50km, file = file.path(data_path_L1,"PAM_plant_site_final_50km.rds"))
saveRDS(PAM_frugivore_site_final_50km, file = file.path(data_path_L1,"PAM_frugivore_site_final_50km.rds"))
```

```{r}
saveRDS(site_loc_key_plant_25km, file = file.path(data_path_L1,"site_loc_key_plant_25km.rds"))
saveRDS(site_loc_key_frugivore_25km, file = file.path(data_path_L1,"site_loc_key_frugivore_25km.rds"))
saveRDS(PAM_plant_site_final_25km, file = file.path(data_path_L1,"PAM_plant_site_final_25km.rds"))
saveRDS(PAM_frugivore_site_final_25km, file = file.path(data_path_L1,"PAM_frugivore_site_final_25km.rds"))
```

```{r}
saveRDS(site_loc_key_plant_10km, file = file.path(data_path_L1,"site_loc_key_plant_10km.rds"))
saveRDS(site_loc_key_frugivore_10km, file = file.path(data_path_L1,"site_loc_key_frugivore_10km.rds"))
saveRDS(PAM_plant_site_final_10km, file = file.path(data_path_L1,"PAM_plant_site_final_10km.rds"))
saveRDS(PAM_frugivore_site_final_10km, file = file.path(data_path_L1,"PAM_frugivore_site_final_10km.rds"))
```

```{r}
saveRDS(site_loc_key_plant_5km, file = file.path(data_path_L1,"site_loc_key_plant_5km.rds"))
saveRDS(site_loc_key_frugivore_5km, file = file.path(data_path_L1,"site_loc_key_frugivore_5km.rds"))
saveRDS(PAM_plant_site_final_5km, file = file.path(data_path_L1,"PAM_plant_site_final_5km.rds"))
saveRDS(PAM_frugivore_site_final_5km, file = file.path(data_path_L1,"PAM_frugivore_site_final_5km.rds"))
```

```{r}
saveRDS(site_loc_key_plant_1km, file = file.path(data_path_L1,"site_loc_key_plant_1km.rds"))
saveRDS(site_loc_key_frugivore_1km, file = file.path(data_path_L1,"site_loc_key_frugivore_1km.rds"))
saveRDS(PAM_plant_site_final_1km, file = file.path(data_path_L1,"PAM_plant_site_final_1km.rds"))
saveRDS(PAM_frugivore_site_final_1km, file = file.path(data_path_L1,"PAM_frugivore_site_final_1km.rds"))
```
